\texttt{treats}: a modular \texttt{R} package for simulating trees and traits.

### 2023-11-22 paper/treats_paper_lineno.tex
Updated paper and response

@@

\title{\texttt{treats}: a modular \texttt{R} package for simulating
trees and traits.}
\author{Thomas Guillerme\(^1\)}
\date{2023-06-09}

@@

Comparing biological patterns is one of the key ways to understand
mechanisms in evolutionary biology. This leads to the development of
phylogenetic comparative methods as key methodologically driven topic in
ecology, evolution and palaeontology (Felsenstein 1985; Pennell and
Harmon 2013). These methods rely on comparing patterns in a phylogenetic
context to understand biological mechanisms or concepts (Harmon 2019).
These comparisons can be done between observed patterns under different
conditions or against null, neutral or baseline models (see Bausman
(2018) for distinctions) suggesting different processes or mechanisms.
For example different traits distribution for species with different
diets (Deepak, Gower, and Cooper 2023) or habitats (Pinto-Ledezma et al.
2017). Or by comparing some observed pattern to one simulated under null
or base conditions (Miller et al. 2022). In theory, workers can follow
the following research pipeline: thinking of a specific mechanism
(e.g.~mass extinction allowing the surviving species to acquire new
morphologies), collecting some data to test this mechanism (e.g.~some
traits of species across and extinction event) and then comparing these
patterns to some simulated under no specific conditions (e.g.~a null
model where the traits evolve randomly regardless of an extinction event
Puttick, Guillerme, and Wills (2020)). Workers might thus need to
simulate a great diversity of evolutionary scenarios to test their
specific question. To do so, we need statistical and software solutions
to simulate trees and data to generate many specific null models.

@@

Although the packages mentioned above are excellent and routinely used
with fast and reliable algorithms and associated documentation, they are
all designed for specific tasks and don't allow much modification beyond
the input parameters designed by the authors. For example,
\texttt{TreeSim} can simulate a birth-death tree with some extinction
event but is not designed to simulated one with an extinction event that
leads to the birth-death process to be not diversity dependent anymore,
simulating a release in selection pressure after the extinction event
that leads to a different process dominating speciation. Or
\texttt{PETER} is not designed to simulate a complex set of traits (say
three correlated BM traits and two independent OU ones). This absence of
modularity has hampered the use of complex and question-driven
simulations, although I acknowledge this was not the primary aim of the
authors of the excellent packages mentioned above. This has led workers
to often develop their own tools to answer specific questions
(e.g.~Puttick, Guillerme, and Wills (2020)). Therefore, I propose
\texttt{treats} a modular \texttt{R} package to simulate both trees and
traits through time.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{3}
\tightlist
\item
  Generating some trait(s) value(s) for the selected lineage (no either
  a tip or a node).
\end{enumerate}

In \texttt{treats}, these three or four steps are implemented as modular
functions that the user can easily change using an internal class of
objects called \texttt{"modifiers"} or \texttt{"traits"} (Fig.
\ref{Fig:workflow}). The simulation then outputs a tree (of class
\texttt{"phylo"} and a associated table of traits - \texttt{"matrix"})
that can be visualised using the \texttt{plot.treats} function. A third
class of object called \texttt{"events"} can be added to the simulations
to modify the entire simulation under certain conditions
(e.g.~simulating a mass extinction). Each element in the algorithm can
be modified by the user using the implemented functions
\texttt{make.bd.params} to set the birth-death parameters,
\texttt{make.traits} to set the trait(s), \texttt{make.modifiers} to set
the birth-death algorithm and \texttt{make.events} to eventually add one
or more events. Users can replace any step in the algorithm by their own
specific functions suiting their needs or, for less advanced users, by
using already implemented functions.

\begin{figure}
\hypertarget{figure1}{%
\centering
\includegraphics{../inst/gitbook/treats_structure.png}
\caption{Figure 1: \texttt{treats} package workflow: the \texttt{treats}
algorithm generates a tree and traits using inbuilt \texttt{"traits"}
objects that contain the instructions on how to generate the trait data
(e.g.~which process? how many dimensions?); \texttt{"modifiers"} objects
that contains instructions on how to ``grow'' the tree (e.g.~by linking
speciation to trait values, or to the current number of species); and
\texttt{"events"} objects that can modify the tree structure,
\texttt{"modifiers"} or \texttt{"traits"} depending on specific
conditions (e.g.~80\% of species with positive trait values go extinct
after reaching a specific time).}\label{figure1}
}
\end{figure}

\hypertarget{brief-applied-example}{%
\subsection{Brief applied example}\label{brief-applied-example}}

To illustrate this we will look whether it is possible to detect changes
in disparity (i.e.~diversity of traits) in a subset of the data
published from Beck and Lee (2014) implemented in Guillerme (2018). This
dataset contains the ordinated traits for 50 mammalian species across
the cretaceous-palaeogene extinction event (K-Pg, 66 Mya; Fig. 2).

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Loading the data and packages}
\KeywordTok{library}\NormalTok{(treats)}
\KeywordTok{library}\NormalTok{(dispRity)}
\KeywordTok{data}\NormalTok{(BeckLee\_tree)}
\KeywordTok{data}\NormalTok{(BeckLee\_mat99)}

\CommentTok{\#\# Creating the time slices}
\NormalTok{time\_slices \textless{}{-}}\StringTok{ }\KeywordTok{chrono.subsets}\NormalTok{(BeckLee\_mat99, BeckLee\_tree,}
                              \DataTypeTok{method =} \StringTok{"continuous"}\NormalTok{,}
                              \DataTypeTok{model  =} \StringTok{"acctran"}\NormalTok{,}
                              \DataTypeTok{time   =} \KeywordTok{seq}\NormalTok{(}\DataTypeTok{from =} \DecValTok{96}\NormalTok{, }\DataTypeTok{to =} \DecValTok{36}\NormalTok{, }\DataTypeTok{by =} \DecValTok{{-}10}\NormalTok{))}

\begin{figure}
\centering
\includegraphics{treats_paper_files/figure-latex/prepdata-1.pdf}
\caption{Figure 2: Observed phylogeny (A) and disparity through time (B)
in the subset of Beck and Lee (2014)'s dataset. Is this change in
disparity related to the K-Pg mass extinction? Or, more generally, is it
even possible to detect potential changes of disparity due to a mass
extinction event?}
\end{figure}

The first and simplest way is to simulate tree topologies that have
similar properties than the observed one. To do so, we need to use some
\textbf{speciation} parameter indicating the rate at which lineages
speciate (\emph{aka} ``birth'' or ``\(\lambda\)'') and an
\textbf{extinction} parameter indicating the rate at which they go
extinct (\emph{aka} ``death'' or ``\(\mu\)''). These two parameters can
be crudely calculated from a given tree by counting the number of
speciations and extinctions through time (but see better methods,
e.g.~\texttt{ape::birthdeath}; Lawton, May, and Raup (1995); Paradis and
Schliep (2019)). We also need a stopping rule for when to stop the
simulations (in our case when reaching 30 time units). This will produce
a single random tree using the input parameters (Fig. 3). Note that I
will not discuss the options here in great details. Much more
information can be found in the
\href{http://tguillerme.github.io/treats.html}{\texttt{treats} manual}.

\CommentTok{\#\# Simulate the tree}
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
\NormalTok{sim\_tree \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{bd.params  =}\NormalTok{ my\_bd\_params,}
                   \DataTypeTok{stop.rule  =}\NormalTok{ stop\_rule,}
                   \DataTypeTok{null.error =} \DecValTok{100}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\centering
\includegraphics{treats_paper_files/figure-latex/simtree-1.pdf}
\caption{Figure 3: A randomly simulated tree with similar properties as
the observed one. Note that the time here is expressed in arbitrary
units.}
\end{figure}

\CommentTok{\#\# Simulate the tree and traits}
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
\NormalTok{sim\_data \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{traits     =}\NormalTok{ my\_traits,}
                   \DataTypeTok{bd.params  =}\NormalTok{ my\_bd\_params,}
                   \DataTypeTok{stop.rule  =}\NormalTok{ stop\_rule,}
                   \DataTypeTok{null.error =} \DecValTok{100}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\centering
\includegraphics{treats_paper_files/figure-latex/simtreetraits-1.pdf}
\caption{Figure 4: A randomly simulated tree with a 2 dimensional random
Brownian Motion trait. Orange, light blue and dark blue dots
respectively represent nodes, fossils and living species.}
\end{figure}

To answer our question, we also want to simulate an extinction event. To
do so, we can create two different \texttt{"events"} object with the
function \texttt{make.events}. The first one will simulate a random
extinction event after reaching half of the simulations (15 time units)
and then making three quarter (0.75) of the taxa go extinct. The second
one will simulate a random extinction but based on trait values: after
reaching time 15, all the species with positive trait values will go
extinct. Both scenarios thus illustrate two different types of mass
extinction (random and selective; Puttick, Guillerme, and Wills (2020)).

\KeywordTok{set.seed}\NormalTok{(}\DecValTok{2}\NormalTok{)}
\CommentTok{\#\# Simulate the tree and traits with a random extinction event}
\NormalTok{sim\_rand\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}
                   \DataTypeTok{traits     =}\NormalTok{ my\_traits,}
                   \DataTypeTok{bd.params  =}\NormalTok{ my\_bd\_params,}
                   \DataTypeTok{stop.rule  =}\NormalTok{ stop\_rule,}
                   \DataTypeTok{events     =}\NormalTok{ random\_extinction,}
                   \DataTypeTok{null.error =} \DecValTok{100}\NormalTok{,}
                   \DataTypeTok{replicates =} \DecValTok{50}\NormalTok{)}

\CommentTok{\#\# Simulate the tree and traits with a random extinction event}
\NormalTok{sim\_trait\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}
                   \DataTypeTok{traits     =}\NormalTok{ my\_traits,}
                   \DataTypeTok{bd.params  =}\NormalTok{ my\_bd\_params,}
                   \DataTypeTok{stop.rule  =}\NormalTok{ stop\_rule,}
                   \DataTypeTok{events     =}\NormalTok{ positive\_extinction,}
                   \DataTypeTok{null.error =} \DecValTok{100}\NormalTok{,}
                   \DataTypeTok{replicates =} \DecValTok{50}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\centering
\includegraphics{treats_paper_files/figure-latex/simtreetraitsevents-1.pdf}
\caption{Figure 5: Two different mass extinction events: (A) 75\% of
species go extinct; (B) all species with positive trait values go
extinct. The red line marks the extinction event.}
\end{figure}

\begin{figure}
\centering
\includegraphics{treats_paper_files/figure-latex/simdisparity-1.pdf}
\caption{Figure 6: Comparison of the observed sum of variances (dashed
line) to the simulated one (ligth and dark grey polygons and black line
representing respectively the 95\% and 50\% confidence interval and the
median value).}
\end{figure}

From these results we can draw some preliminary conclusions: the
observed change in disparity is more likely due to a random mass
extinction rather than a selective one. This is of course a very crude
way of testing this, a more rigorous approach is needed to answer the
question: more and better quality data, and more thorough methods
(e.g.~using a rank envelope test Murrell (2018)).

@@

Traits are simulated via the \texttt{make.traits} function given one or
more processes, number of dimensions per process and some starting
value(s). Essentially, the generation of new trait values is based on a
process (a \texttt{function}) modifying a trait value (\texttt{x0})
relative to some branch length (\texttt{edge.length}). For example, a
simple Brownian Motion process could be generated by the function
\texttt{rnorm} where the it modifies the trait value \texttt{x0}
relative to the branch length (\texttt{edge.length}). The longer the
branch length, the more likely the new trait value will be different
from \texttt{x0}:

@@

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Creating an extinction that removes species with positive trait values}
\NormalTok{positive\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
    \DataTypeTok{target       =} \StringTok{"taxa"}\NormalTok{,}
    \DataTypeTok{condition    =} \KeywordTok{time.condition}\NormalTok{(}\DecValTok{15}\NormalTok{),}
    \DataTypeTok{modification =} \KeywordTok{random.extinction}\NormalTok{(}\FloatTok{0.75}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

>>>>

\usepackage{lineno}

@@

\linenumbers

\title{\texttt{treats}: a modular \texttt{R} package for simulating
trees and traits.}
\author{Thomas Guillerme\(^1\)}
\date{2023-11-22}

@@

Comparing biological patterns is one of the key ways to understand
mechanisms in evolutionary biology. This leads to the development of
phylogenetic comparative methods as key methodologically driven topic in
ecology, evolution and palaeontology (Felsenstein 1985; Pennell and
Harmon 2013). These methods rely on comparing patterns in a phylogenetic
context to understand biological mechanisms or concepts (Harmon 2019).
These comparisons can be done between observed patterns under different
conditions or against null, neutral or baseline models (see Bausman
(2018) for distinctions) suggesting different processes or mechanisms.
For example different traits distribution for species with different
diets (Deepak, Gower, and Cooper 2023) or habitats (Pinto-Ledezma et al.
2017). Or by comparing some observed pattern to one simulated under null
or base conditions (Miller et al. 2022). In theory, workers can follow
the following research pipeline: 1) thinking of a specific mechanism
(e.g.~mass extinction allowing the surviving species to acquire new
morphologies), 2) collecting some data to test this mechanism (e.g.~some
traits of species across and extinction event) and then 3) comparing
these patterns to some simulated under no specific conditions (e.g.~a
null model where the traits evolve randomly regardless of an extinction
event Puttick, Guillerme, and Wills (2020)). Workers might thus need to
simulate a great diversity of evolutionary scenarios to test their
specific question. To do so, we need statistical and software solutions
to simulate trees and data to generate many specific null models.

@@

Although the packages mentioned above are excellent and routinely used
with fast and reliable algorithms and associated documentation, they are
all designed for specific tasks and don't allow much modification beyond
the input parameters designed by the authors. For example,
\texttt{TreeSim} can simulate a birth-death tree with some extinction
event but is not designed to simulated one with an extinction event that
leads to the birth-death process to be not diversity dependent anymore,
simulating a release in selection pressure after the extinction event
that leads to a different process dominating speciation. Or
\texttt{PETER} is not designed to simulate a complex set of traits (say
three correlated BM traits and two independent OU ones). This absence of
modularity has hampered the use of complex and question-driven
simulations, although I acknowledge this was not the primary aim of the
authors of the excellent packages mentioned above. This has led workers
to often develop their own tools to answer specific questions
(e.g.~Puttick, Guillerme, and Wills (2020)). Therefore, I propose
\texttt{treats} a modular \texttt{R} package to simulate both trees and
traits through time. Note that although \texttt{treats} is modular and
thus allows to be used as go to tool for simulating and trees and
traits, it lacks the ready-to-use implemented methods featured in other
packages such as fossilisation and sampling (Barido-Sottani et al. 2019;
Stadler 2011; Rosario Petrucci, Januario, and Quental 2022) or specific
macroevolutionary simulations (Morlon et al. 2016; Puttick, Guillerme,
and Wills 2020).

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{3}
\tightlist
\item
  Generating some trait(s) value(s) for the selected lineage (either a
  tip or a node - but see the ``Simulating traits section'' below to
  generate trait values along edges).
\end{enumerate}

In \texttt{treats}, these three or four steps are implemented as modular
functions that the user can easily change using an internal class of
objects called \texttt{"modifiers"} or \texttt{"traits"} (Fig. 1). The
simulation then outputs a tree (of class \texttt{"phylo"} and a
associated table of traits - \texttt{"matrix"}) that can be visualised
using the \texttt{plot.treats} function. A third class of object called
\texttt{"events"} can be added to the simulations to modify the entire
simulation under certain conditions (e.g.~simulating a mass extinction).
Each element in the algorithm can be modified by the user using the
implemented functions \texttt{make.bd.params} to set the birth-death
parameters, \texttt{make.traits} to set the trait(s),
\texttt{make.modifiers} to set the birth-death algorithm and
\texttt{make.events} to eventually add one or more events. Users can
replace any step in the algorithm by their own specific functions
suiting their needs or, for less advanced users, by using already
implemented functions.

\begin{figure}
\hypertarget{figure1}{%
\centering
\includegraphics{../inst/gitbook/treats_structure.png}
\caption{\texttt{treats} package workflow: the \texttt{treats} algorithm
generates a tree and traits using inbuilt \texttt{"traits"} objects that
contain the instructions on how to generate the trait data (e.g.~which
process? how many dimensions?); \texttt{"modifiers"} objects that
contains instructions on how to ``grow'' the tree (e.g.~by linking
speciation to trait values, or to the current number of species); and
\texttt{"events"} objects that can modify the tree structure,
\texttt{"modifiers"} or \texttt{"traits"} depending on specific
conditions (e.g.~80\% of species with positive trait values go extinct
after reaching a specific time).}\label{figure1}
}
\end{figure}

@@

Traits are simulated via the \texttt{make.traits} function given one or
more processes, number of dimensions per process and some starting
value(s). Essentially, the generation of new trait values is based on a
process (a \texttt{function}) modifying a trait value (\texttt{x0})
relative to some branch length (\texttt{edge.length}). For example, a
simple Brownian Motion process could be generated by the function
\texttt{rnorm} where it modifies the trait value \texttt{x0} relative to
the branch length (\texttt{edge.length}). The longer the branch length,
the more likely the new trait value will be different from \texttt{x0}:

@@

Traits are always generated (and stored) only for tips or nodes and not
along edges. However, it is possible so generate trait values at
specific or regular time intervals by using the option
\texttt{save.steps} in the \texttt{treats} function. This will generate
singleton nodes (i.e.~nodes with only one descendant) with associated
trait values.

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Creating an extinction that removes species with positive trait values}
\NormalTok{positive\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
    \DataTypeTok{target       =} \StringTok{"taxa"}\NormalTok{,}
    \DataTypeTok{condition    =} \KeywordTok{time.condition}\NormalTok{(}\DecValTok{15}\NormalTok{),}
    \DataTypeTok{modification =} \KeywordTok{trait.extinction}\NormalTok{(}\DataTypeTok{x =} \DecValTok{0}\NormalTok{, }\DataTypeTok{condition =} \StringTok{\textasciigrave{}}\DataTypeTok{\textgreater{}=}\StringTok{\textasciigrave{}}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

\hypertarget{brief-applied-example}{%
\section{Brief applied example}\label{brief-applied-example}}

To illustrate this we will look whether it is possible to detect changes
in disparity (i.e.~diversity of traits) in a subset of the data
published from Beck and Lee (2014) implemented in Guillerme (2018). This
dataset contains the ordinated traits for 50 mammalian species across
the Cretaceous-Palaeogene extinction event (K-Pg, 66 Mya; Fig. 2).

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Loading the package and data}
\KeywordTok{library}\NormalTok{(treats)}
\KeywordTok{data}\NormalTok{(BeckLee\_tree)}
\KeywordTok{data}\NormalTok{(BeckLee\_mat99)}

\CommentTok{\#\# Creating the time slices}
\NormalTok{time\_slices \textless{}{-}}\StringTok{ }\KeywordTok{chrono.subsets}\NormalTok{(BeckLee\_mat99, BeckLee\_tree,}
                              \DataTypeTok{method =} \StringTok{"continuous"}\NormalTok{,}
                              \DataTypeTok{model  =} \StringTok{"proximity"}\NormalTok{,}
                              \DataTypeTok{time   =} \KeywordTok{seq}\NormalTok{(}\DataTypeTok{from =} \DecValTok{96}\NormalTok{, }\DataTypeTok{to =} \DecValTok{36}\NormalTok{, }\DataTypeTok{by =} \DecValTok{{-}10}\NormalTok{))}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Plotting the tree and the disparity through time}
\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{), }\DataTypeTok{bty =} \StringTok{"null"}\NormalTok{)}
\CommentTok{\#\# Thre tree}
\KeywordTok{plot}\NormalTok{(BeckLee\_tree, }\DataTypeTok{show.tip.label =} \OtherTok{FALSE}\NormalTok{)}
\KeywordTok{axisPhylo}\NormalTok{()}
\KeywordTok{abline}\NormalTok{(}\DataTypeTok{v =}\NormalTok{ BeckLee\_tree}\OperatorTok{$}\NormalTok{root.time }\OperatorTok{{-}}\StringTok{ }\DecValTok{66}\NormalTok{, }\DataTypeTok{col =} \StringTok{"red"}\NormalTok{)}
\KeywordTok{legend}\NormalTok{(}\StringTok{"topleft"}\NormalTok{, }\DataTypeTok{pch =} \OtherTok{NULL}\NormalTok{, }\DataTypeTok{legend =} \StringTok{"A"}\NormalTok{, }\DataTypeTok{bty =} \StringTok{"n"}\NormalTok{, }\DataTypeTok{cex =} \DecValTok{2}\NormalTok{)}

\CommentTok{\#\# The disparity}
\KeywordTok{plot}\NormalTok{(observed\_disparity, }\DataTypeTok{ylab =} \StringTok{"Sum of variances"}\NormalTok{)}
\KeywordTok{abline}\NormalTok{(}\DataTypeTok{v =} \DecValTok{4}\NormalTok{, }\DataTypeTok{col =} \StringTok{"red"}\NormalTok{)}
\KeywordTok{legend}\NormalTok{(}\StringTok{"topleft"}\NormalTok{, }\DataTypeTok{pch =} \OtherTok{NULL}\NormalTok{, }\DataTypeTok{legend =} \StringTok{"B"}\NormalTok{, }\DataTypeTok{bty =} \StringTok{"n"}\NormalTok{, }\DataTypeTok{cex =} \DecValTok{2}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\centering
\includegraphics{treats_paper_files/figure-latex/prepdata-1.pdf}
\caption{Observed phylogeny (A) and disparity through time (B) in the
subset of Beck and Lee (2014)'s dataset. The red line represents the
K-Pg boundary. Is this change in disparity related to the K-Pg mass
extinction? Or, more generally, is it even possible to detect potential
changes of disparity due to a mass extinction event?}
\end{figure}

The first and simplest way is to simulate tree topologies that have
similar properties than the observed one. To do so, we need to use some
\textbf{speciation} parameter indicating the rate at which lineages
speciate (\emph{aka} ``birth'' or ``\(\lambda\)'') and an
\textbf{extinction} parameter indicating the rate at which they go
extinct (\emph{aka} ``death'' or ``\(\mu\)''). These two parameters can
be crudely calculated from a given tree by counting the number of
speciations and extinctions through time (but see better methods,
e.g.~\texttt{ape::birthdeath}; Lawton, May, and Raup (1995); Paradis and
Schliep (2019)). We also need a stopping rule for when to stop the
simulations (in our case when reaching 30 time units). This will produce
a single random tree using the input parameters (Fig. 3). The number of
time units in \texttt{treats} is arbitrary and is not equivalent to
millions of years. Using 30 time units here allows to simulate number of
tips in a similar order of magnitude as the ones in the observed data.
Note that I will not discuss the options here in great details. Much
more information can be found in the
\href{http://tguillerme.github.io/treats.html}{\texttt{treats} manual}.

\CommentTok{\#\# Simulate the tree}
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
\NormalTok{sim\_tree \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{bd.params  =}\NormalTok{ my\_bd\_params,}
                   \DataTypeTok{stop.rule  =}\NormalTok{ stop\_rule,}
                   \DataTypeTok{null.error =} \DecValTok{100}\NormalTok{,}
                   \DataTypeTok{verbose    =} \OtherTok{FALSE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\centering
\includegraphics{treats_paper_files/figure-latex/simtree-1.pdf}
\caption{A randomly simulated tree with similar properties as the
observed one. Note that the time here is expressed in arbitrary units.}
\end{figure}

\CommentTok{\#\# Simulate the tree and traits}
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
\NormalTok{sim\_data \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{traits     =}\NormalTok{ my\_traits,}
                   \DataTypeTok{bd.params  =}\NormalTok{ my\_bd\_params,}
                   \DataTypeTok{stop.rule  =}\NormalTok{ stop\_rule,}
                   \DataTypeTok{null.error =} \DecValTok{100}\NormalTok{,}
                   \DataTypeTok{verbose    =} \OtherTok{FALSE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
\CommentTok{\#\# Plotting one trait through time}
\KeywordTok{plot}\NormalTok{(sim\_data, }\DataTypeTok{ylab =} \StringTok{"Trait 1"}\NormalTok{, }\DataTypeTok{las =} \DecValTok{1}\NormalTok{, }\DataTypeTok{main =} \StringTok{"Trait 1 through time"}\NormalTok{)}

\CommentTok{\#\# Plotting the two dimensions against each other}
\KeywordTok{plot}\NormalTok{(sim\_data, }\DataTypeTok{trait =} \KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{1}\NormalTok{), }\DataTypeTok{ylab =} \StringTok{"Trait 1"}\NormalTok{, }\DataTypeTok{xlab =} \StringTok{"Trait 2"}\NormalTok{, }\DataTypeTok{las =} \DecValTok{1}\NormalTok{, }\DataTypeTok{main =} \StringTok{"Traits 1 and 2"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\centering
\includegraphics{treats_paper_files/figure-latex/simtreetraits-1.pdf}
\caption{A randomly simulated tree with a 2 dimensional random Brownian
Motion trait. Orange, light blue and dark blue dots respectively
represent nodes, fossils and living species.}
\end{figure}

To answer our question, we also want to simulate an extinction event. To
do so, we can create two different \texttt{"events"} object with the
function \texttt{make.events}. The first one will simulate a random
extinction event after reaching half of the simulations (15 time units)
and then making three quarter (0.75) of the taxa go extinct. The second
one will simulate a random extinction but based on trait values: after
reaching time 15, all the species with positive trait values will go
extinct. Both scenarios illustrate two different types of mass
extinctions but they are not equivalent: because of the ancestral trait
value starting at 0, we expect the second scenario to remove on average
only 50\% of the species (i.e.~half the species are expected to evolve a
trait value above 0). For more details on simulating the effect of mass
extinction and the difficulties to simulate an unambiguous effect of a
mass extinction, see Puttick, Guillerme, and Wills (2020). Because of
this stochasticity of the simulations, we will repeat them 50 times
(using \texttt{replicates\ =\ 50}) to generate a distribution of
possible simulated scenarios as opposed to a random single one that
could be idiosyncratic (Fig. 5).

\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
\CommentTok{\#\# Simulate the tree and traits with a random extinction event}
\NormalTok{sim\_rand\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}
                   \DataTypeTok{traits     =}\NormalTok{ my\_traits,}
                   \DataTypeTok{bd.params  =}\NormalTok{ my\_bd\_params,}
                   \DataTypeTok{stop.rule  =}\NormalTok{ stop\_rule,}
                   \DataTypeTok{events     =}\NormalTok{ random\_extinction,}
                   \DataTypeTok{null.error =} \DecValTok{100}\NormalTok{,}
                   \DataTypeTok{replicates =} \DecValTok{50}\NormalTok{)}

\CommentTok{\#\# Simulate the tree and traits with a selective extinction event}
\NormalTok{sim\_trait\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}
                   \DataTypeTok{traits     =}\NormalTok{ my\_traits,}
                   \DataTypeTok{bd.params  =}\NormalTok{ my\_bd\_params,}
                   \DataTypeTok{stop.rule  =}\NormalTok{ stop\_rule,}
                   \DataTypeTok{events     =}\NormalTok{ positive\_extinction,}
                   \DataTypeTok{null.error =} \DecValTok{100}\NormalTok{,}
                   \DataTypeTok{replicates =} \DecValTok{50}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\centering
\includegraphics{treats_paper_files/figure-latex/simtreetraitsevents-1.pdf}
\caption{Two different mass extinction events: (A) 75\% of species go
extinct; (B) all species with positive trait values go extinct. The red
line marks the extinction event.}
\end{figure}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Remove single nodes simulated at the extinction}
\NormalTok{sim\_rand\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{drop.singles}\NormalTok{(sim\_rand\_extinction)}
\NormalTok{sim\_trait\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{drop.singles}\NormalTok{(sim\_trait\_extinction)}

\CommentTok{\#\# Calculate the dispRity for all the simulations}
\NormalTok{random\_extinction\_disparity \textless{}{-}}\StringTok{ }\KeywordTok{dispRitreats}\NormalTok{(sim\_rand\_extinction,}
                                            \DataTypeTok{method =} \StringTok{"continuous"}\NormalTok{,}
                                            \DataTypeTok{model  =} \StringTok{"proximity"}\NormalTok{,}
                                            \DataTypeTok{time   =} \KeywordTok{seq}\NormalTok{(}\DataTypeTok{from =} \DecValTok{25}\NormalTok{, }\DataTypeTok{to =} \DecValTok{0}\NormalTok{, }\DataTypeTok{by =} \DecValTok{{-}5}\NormalTok{),}
                                            \DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sum, variances),}
                                            \DataTypeTok{scale.trees =} \OtherTok{FALSE}\NormalTok{)}
\NormalTok{selective\_extinction\_disparity \textless{}{-}}\StringTok{ }\KeywordTok{dispRitreats}\NormalTok{(sim\_trait\_extinction,}
                                            \DataTypeTok{method =} \StringTok{"continuous"}\NormalTok{,}
                                            \DataTypeTok{model  =} \StringTok{"proximity"}\NormalTok{,}
                                            \DataTypeTok{time   =} \KeywordTok{seq}\NormalTok{(}\DataTypeTok{from =} \DecValTok{25}\NormalTok{, }\DataTypeTok{to =} \DecValTok{0}\NormalTok{, }\DataTypeTok{by =} \DecValTok{{-}5}\NormalTok{),}
                                            \DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sum, variances),}
                                            \DataTypeTok{scale.trees =} \OtherTok{FALSE}\NormalTok{)}

\CommentTok{\#\# Scale the disparity results (to compare to the observed ones)}
\NormalTok{random\_extinction\_disparity \textless{}{-}}\StringTok{ }\KeywordTok{rescale.dispRity}\NormalTok{(random\_extinction\_disparity)}
\NormalTok{selective\_extinction\_disparity \textless{}{-}}\StringTok{ }\KeywordTok{rescale.dispRity}\NormalTok{(selective\_extinction\_disparity)}
\NormalTok{observed\_disparity \textless{}{-}}\StringTok{ }\KeywordTok{unlist}\NormalTok{(}\KeywordTok{get.disparity}\NormalTok{(}\KeywordTok{rescale.dispRity}\NormalTok{(observed\_disparity)))}

\CommentTok{\#\# Plotting the results with the observed disparity}
\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
\KeywordTok{plot}\NormalTok{(random\_extinction\_disparity, }\DataTypeTok{main =} \StringTok{"Random extinction"}\NormalTok{, }\DataTypeTok{ylab =} \StringTok{"Scaled sum of variances"}\NormalTok{)}
\KeywordTok{abline}\NormalTok{(}\DataTypeTok{v =} \DecValTok{3}\NormalTok{, }\DataTypeTok{col =} \StringTok{"red"}\NormalTok{)}
\KeywordTok{lines}\NormalTok{(}\DataTypeTok{x =} \DecValTok{1}\OperatorTok{:}\DecValTok{7}\NormalTok{, }\DataTypeTok{y =}\NormalTok{ observed\_disparity, }\DataTypeTok{lty =} \DecValTok{2}\NormalTok{, }\DataTypeTok{lwd =} \DecValTok{2}\NormalTok{)}
\KeywordTok{plot}\NormalTok{(selective\_extinction\_disparity, }\DataTypeTok{main =} \StringTok{"Selective extinction"}\NormalTok{, }\DataTypeTok{ylab =} \StringTok{""}\NormalTok{)}
\KeywordTok{abline}\NormalTok{(}\DataTypeTok{v =} \DecValTok{3}\NormalTok{, }\DataTypeTok{col =} \StringTok{"red"}\NormalTok{)}
\KeywordTok{lines}\NormalTok{(}\DataTypeTok{x =} \DecValTok{1}\OperatorTok{:}\DecValTok{7}\NormalTok{, }\DataTypeTok{y =}\NormalTok{ observed\_disparity, }\DataTypeTok{lty =} \DecValTok{2}\NormalTok{, }\DataTypeTok{lwd =} \DecValTok{2}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\centering
\includegraphics{treats_paper_files/figure-latex/simdisparity-1.pdf}
\caption{Comparison of the observed sum of variances (dashed line) to
the simulated one (ligth and dark grey polygons and black line
representing respectively the 95\% and 50\% confidence interval and the
median value).}
\end{figure}

From these results we can draw some preliminary conclusions: the
observed change in disparity is more likely due to a selective mass
extinction rather than a random one (Fig. 6). This is of course a very
crude way of testing this, a more rigorous approach is needed to answer
the question: more and better quality data, and more thorough methods
(e.g.~using a rank envelope test Murrell (2018)).
